Fail Faster

### 2025-10-06 paper/paper.tex
draft 3.3

\subsection{Staging Combinators}

In Section~\ref{section:motiv}, we noted that generator combinators like \texttt{weighted\_union}
allocate lists in the hot path of the generator. Even though these lists are usually small---at most a few dozen elements---each allocation takes us closer to the next garbage collection, which is bad for performance.

This is an ideal opportunity to exercise another feature of staging: compile-time specialization.
Since we almost always know the particular list of choices at compile time, a staged version of \texttt{weighted\_union}
can generate \emph{different code} depending on the number of generators in the union.
If we use weighted union on a compile-time list of generators \texttt{g1}, \texttt{g2}, and \texttt{g3},
we can emit code that picks between the generators without realizing the list at
run time.

@@

Figure~\ref{fig:staged-weighted-union} shows the code for such a staged weighted union.
Crucially, it takes a \emph{compile-time} list \texttt{weighted\_gens} of
generators and weights. The weights themselves might only be known at run time---it is
common to use the current size parameter as a weight, for instance---so they are \texttt{code}s.
Instead of building a data structure representing a histogram of the distribution described by the weights at run time and then traversing it,
the compile-time \texttt{weighted\_union} combinator generates a
tree of \texttt{if}s, specialized to the list of weights known at compile time, that picks out the selected generator.

>>>>

\subsection{PBT-Specific Staging}
While the techniques in the previous section eliminates monadic
DSL overhead using standard ideas in staging, it cannot address some of the
more subtle and domain-specific performance issues identified in Section~\ref{section:motiv},
like hot-path allocations induced by generator combinators such as \texttt{weighted\_union}.
Removing these allocations requires a observation that transcends simple 
inlining: though the \emph{elements} of the lists may be dynamic, their
\emph{spine}---the structure of the list itself---is in practice always statically
known. This is an instance of \emph{partially-static data} \cite{partially-static}:
a data structure whose shape is static but whose contents are dynamic.

This observation enables an optimization unique to the PBT generator library setting.
Because the spine is static, we can unfold the choice-selection
logic over the list at compile time time, specializing \texttt{weighted\_union} to the exact number
and weights of choices. The result is a specialized generator that branches directly to the chosen
alternative, without ever materializing the list of choices at run time.

% In Section~\ref{section:motiv}, we noted that generator combinators like
% \texttt{weighted\_union} allocate lists in the hot path of the generator. Even
% though these lists are usually small---at most a few dozen elements---each
% allocation takes us closer to the next garbage collection, which is bad for
% performance.

% The simple staging we've used thusfar to remove the overhead of \texttt{return}
% and \texttt{bind} cannot eliminate this allocation --- we must use structure specific
% to PBT generator combinators to our advantage.

% In particular, the \emph{spine} of the list of choices that \texttt{weighted\_union} picks from is
% in practice always statically known, even if the elements themselves may be determined dynamically
% --- the list is \emph{partially-static data} \cite{partially-static}.
% Because of this, we can statically unroll the outer loop of the \texttt{weighted\_union} combinator
% and specialize to the potential choices at compile-time.

% This is an ideal opportunity to exercise another feature of staging: compile-time specialization.

% Since we almost always know the particular list of choices at compile time, a staged version of \texttt{weighted\_union}
% can generate \emph{different code} depending on the number of generators in the union.
% If we use weighted union on a compile-time list of generators \texttt{g1}, \texttt{g2}, and \texttt{g3},
% we can emit code that picks between the generators without realizing the list at
% run time.

@@

Figure~\ref{fig:staged-weighted-union} shows the code for such a staged weighted union.
Crucially, it takes a \emph{compile-time} list \texttt{weighted\_gens} of
generators and weights. The weights themselves might only be known at run time---it is
common to use the current size parameter as a weight, for instance---so they are \texttt{code}s.
Instead of building a data structure representing a histogram of the distribution described by the weights at run time and then traversing it,
the compile-time \texttt{weighted\_union} combinator generates a
tree of \texttt{if}s that picks out the selected generator.
